Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-05 talk/dls2014/paper/paper.tex
adapt the style of writing

A naive approach to providing this complete isolation between threads
is to partition the virtual memory of a process into $N$ segments, one
for each thread currently executing.  Each segment then holds a
complete copy of the memory logically seen by the program.  All
threads read and write data inside their own segment; exchanges
between segments occur only when transactions commit.  So far, this is a
model that would also work for distributed transactional memory.

@@

When we need to store into our objects pointers to other objects, we
need to be careful: if we were to directly store regular addresses valid
in the local segment, then the objects' binary content would have to differ
between segments.  This would prevent any sharing.  Instead, we store
pointers as \emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start
of the segment.  A $SO$ can then be interpreted from any segment by
adding it to the base address of that segment.\footnote{An alternative
model would be to simulate threads using several independent processes
in which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the state
(file descriptors, external libraries, ...) is shared.} The result is
called a \emph{Linear Address (LA)}. This is illustrated in
Figure~\ref{fig:Segment-Addressing}.

The x86 CPUs provide a feature which is precisely called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment register
$\%gs$, which is available to applications, to point to the
thread's segment start address.\footnote{The other segment register
$\%fs$ is typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for a
similar purpose: thread-local data -- but a lot of it.}  The translation
from a $SO$ to a LA is done for us by using the ``$\%gs\colon$'' prefix
before any CPU instruction that reads or writes memory using a $SO$
address.  This is very efficient and some compilers support it natively
(e.g.\ clang).

If support for this is not available (due to the compiler or on a
different architecture), there are other optimizations possible.  For
example, using gcc, a regular register can be globally reserved and
all ``simple enough'' memory accesses will still resolve to a single
instruction.  More complicated accesses, e.g.\ those involving array
indexing, will need an extra explicit addition.

@@

What we described so far has no mesurable performance impact by
itself, and it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects in the read or write set before
reading or writing to them\footnote{Barriers added by an automatic and
local program transformation~\cite{felber07}}.  Furthermore, we need
to describe the commit protocol.

@@

\item [{Read~Barrier:}] A major contribution to our performance results
  is the simplicity of the read barrier: a large program (like an
  interpreter for a complex language) contains a number of reads that
  typically exceeds many times the number of writes.  The described
  approach allows the read barrier to be a single, unconditional memory
  write into segment-local memory.
 We managed to
  design the read barrier so that it is done by one memory write
  (setting a flag) into non-shared memory.  This is done with a
  segment-local array of flags indexed by a number derived from the $SO$
  of the object.  Unlike other STM systems, the read barrier does not
  have to find the private version of some object -- our two-step
  address translation automatically resolves the reference to the
  private version on every access anyway.  This means the read barrier
  is not constrained to execute before the actual read -- both the
  compiler and the CPU are free to reorder them.

\item [{Write~Barrier:}] This is triggered by checking a flag in the
  header of the object.  When the flag is set, we call the slow path,
  which adds the object into the write set and unconditionally resets
  the flag.  We eagerly detect write-write conflicts by allowing only
  one transaction modifying an object at a time (by acquiring a write
  lock).\footnote{Eager write-write detection is not inherent to our
  approach; future experiments may show that we want to lift this
  restriction.}  This slow path is also responsible for privatising
  pages,\footnote{The write set contains objects, not pages.} as
  described above.  Interestingly, we are using a generational garbage
  collector (described below) which requires a similar write barrier
  anyway -- so the same header flag is used for both cases.  The flag
  means ``something needs to be done before writing to this object'',
  and the slow path logic determines what it is.

\item [{Commit:}] When we want to commit a transaction, we need to check
  for read-write conflicts.  Such conflicts cannot be detected in the
  minimal read barrier, but can be detected during commit: the case to
  look for is that the committing transaction wrote to an object that
  another transaction has already read.\footnote{Which is why the read
  sets can be just sparsely populated arrays of flags: we never walk the
  whole read set.} A priori, checking other transactions' read sets
  would involve reading concurrently modified memory: a different thread
  could be issuing a read barrier in parallel with an unprotected
  commit.  To prevent this from occurring, we must suspend all other
  threads at known safe points.  This is a trade-off for the simplicity
  of our read barriers -- which is well worth it in our case, due to the
  very high number of read barriers when compared to the number of
  commits and the (low) total number of threads.

@@

Conflicts can be resolved by aborting one of the two transactions, or in
some cases by pausing one of them until the other commits.  If a commit
succeeds, we copy the changes to modified objects so that they become
visible in other segments.  Note that this only includes objects that
existed before the transaction and have been modified by it.  It does
not include all the new objects created by the current transaction
(there is typically a very high rate of object churn in dynamic languages).
Our garbage collector allocates new objects in pages that are shared in
the first place, so that they do not need to be copied -- and moreover,
such objects will commonly never be modified any more.

Privatised pages are re-shared during major collection cycles.

>>>>

A naive approach to providing this complete isolation between threads
is to partition the virtual memory of a process into $N$ segments, one
for each thread currently executing.  Each segment then holds a
complete copy of the memory logically seen by the program.  All
threads read and write data inside their own segment; and this data
gets copied to other segments in an atomic fashion only when a
transaction commits.  So far, this is a model that would also work for
distributed transactional memory.

@@

When we store into our objects the references to other objects, we
need to be careful: addresses that are valid in one segment are not
valid in another, because they would break isolation by pointing to
the copy of the object in another segment.  Consequently, we would
need to adapt these addresses to the segment they live in, which would
prevent completely any page sharing.  Instead, we store pointers as
\emph{Segment Offsets ($SO$)}, i.e.\ offsets from the start of the
segment.  A $SO$ can then be interpreted in any segment by adding to
it the start address of that segment.\footnote{An alternative model
would be to simulate threads using several independent processes in
which all segments are at the same address.  This might work well for
demonstration purposes, but it is not really suitable for implementing
an existing programming language: programs assume that most of the
state (file descriptors, external libraries, ...) is shared.} The
result is called a \emph{Linear Address (LA)}. This is illustrated in
Figure~\ref{fig:Segment-Addressing}.

The x86 CPUs provide a feature which is called \emph{memory
segmentation}.  On the modern x86-64, this is enough to perform the
addition described above very efficiently.  We use the segment
register $\%gs$\footnote{The other segment register $\%fs$ is
typically used by the threading library to provide access to
thread-local data.  One point of view is that we are using $\%gs$ for
a similar purpose: thread-local data -- but a lot of it.}, which is
available to applications, to point to the current thread's segment
start address.  The translation from a $SO$ to a LA is done for us by
using the ``$\%gs\colon$'' prefix before any CPU instruction that
reads or writes memory using a $SO$ address.  This is very efficient
(we can do it on every object access) and some compilers support it
natively (e.g.\ clang).

In summary, translating a $\%gs:SO$ to a physical address is a
two-step process: First the memory segmentation feature of the CPU
constructs a linear address. Then, this LA gets mapped by the MMU to
some place in the physical memory. This makes the $SO$ a valid reference
to an object in all segments.

% If support for this is not available (due to the compiler or a
% different CPU architecture), there are other optimisations possible.
% For example, using gcc, a regular register can be globally reserved
% and all ``simple enough'' memory accesses will still resolve to a
% single instruction.  More complicated accesses, e.g.\ those involving
% array indexing, will need an extra explicit addition.

@@

What we described so far has no measurable performance overhead by
itself, but it is missing conflict detection to be considered a
full-fledged STM system.  This requires adding \emph{barriers} to the
program to register the objects in the read or write set before
reading or writing to them\footnote{Barriers added by an automatic and
local program transformation~\cite{felber07}}.  Furthermore, we still
need to describe the commit protocol.

@@

\item [{Read~Barrier:}] A major contribution to our performance
  is the simplicity of the read barrier: a large program (like an
  interpreter for a complex language) contains a number of reads that
  typically exceeds many times the number of writes.  The described
  approach allows the read barrier to be a single, unconditional
  write into segment-local memory. We only need to mark an object
  as read by the current transaction. The mark is stored in an
  segment-local array indexed by a number derived from the $SO$
  of the object.

  Unlike other STM systems, the read barrier does not have to find the
  private version of the object -- our two-step address translation
  automatically resolves the reference to the private version on every
  access anyway\footnote{Additional benefit: The read barrier is not
    constrained to execute before the actual read -- both the compiler and
    the CPU are free to reorder them.}.

\item [{Write~Barrier:}] This barrier is triggered by checking a flag
  on the object that signifies if we did not already execute the write
  barrier on this object before.  When the flag is set, we call the slow
  path, which adds the object into the write set and unconditionally
  resets the flag.

  Here, we eagerly detect write-write conflicts by allowing only one
  transaction modifying an object at a time (by acquiring a write
  lock).\footnote{Eager write-write detection is not inherent to our
    approach; we may lift this restriction in the future.}  Part of making
  the object write-ready is also the privatisation of all pages the
  object belongs to as described above.

  Described below, we also need a write barrier for our generational
  garbage collector (GC). Interestingly, we are able to share the same
  barrier and flag for both use cases. Since the GC is an integral
  part of many dynamic languages, we actually do not incur any additional
  overhead here since the barriers are already there.

\item [{Commit:}] When we want to commit a transaction, we need to check
  for read-write conflicts.  Such conflicts cannot be detected in the
  minimal read barrier, but can be detected during commit: we check
  if any of the objects in our write set was already read by another
  transaction. We never walk the read set, but just check if the objects
  are marked in other segments.

  Checking other transactions' read marks would involve reading
  concurrently modified memory: another thread could be issuing a read
  barrier in parallel with a commit.  To prevent this from occurring, we
  must suspend all other threads at known safe points.  This is a
  trade-off for the simplicity of our read barrier -- which is well
  worth it in our case, due to the very high number of read barriers
  when compared to the number of commits.

@@

Conflicts can be resolved by aborting one of the two transactions, or
in some cases by pausing one of them until the other commits.  If a
commit succeeds, we copy the changes made to the objects in the write
set so that they become visible in other segments.  Note that this
only includes objects that existed before the transaction and have
been modified by it.  It does not include all the new objects created
by the current transaction, which typically outnumber modified objects
by far in dynamic languages.  We allocate these new objects in pages
that are still shared, so that they do not need to be copied to other
segments at all.
